### 2021-04-09 introduction.tex
a pass on abstract and intro

\begin{document}
\section{Introduction}
Session types are a type formalism used to specify and verify communication protocols between two or more communicating agents~\cite{honda93,takeuchihonda94,hondavasconcelos98,carbonehonda07}.
They have been studied for many programming paradigms, including concurrent and functional paradigms. In particular, they have been defined for the $\pi$-calculus~\cite{sangiorgiwalker01}---a process calculus for communication and concurrency, and for Good Variation (GV)~\cite{wadler14,lindleymorris15}---a linear concurrent $\lambda$-calculus.
Session type systems guarantee communication safety and session fidelity, namely that communication proceeds without any type mismatch and it follows the predefined protocol specified as a session type.
Another key property in a concurrent setting is deadlock freedom, stating that agents communicate without getting stuck in cyclic dependencies. Deadlock freedom is a more involving property and session types alone are not enough to rule out deadlocks in cyclic structures of multiple interleaved sessions. Several techniques have been developed to address deadlock freedom in the $\pi$-calculus and concurrent $\lambda$-calculus with session types, as detailed below.

With regards to $\pi$-calculus, a growing line of work on deadlock freedom leverages the Curry-Howard correspondences between intuitionistic or classical linear logic and $\pi$-calculus with session types~\cite{cairespfenning10,wadler14}, which guarantee deadlock freedom by design. This is achieved by merging constructs for parallel composition and channel restriction under the logical cut rule, which forces processes to share only one channel for communication and eliminates cyclic dependencies. Recent developments in the Classical Processes (CP)~\cite{wadler14} line of work have led to various approaches to decoupling these constructs, either by maintaining a strong correspondence to classical linear logic, as in Hypersequent CP~\cite{kokkemontesi19popl,kokkemontesi19tlla}, or by weakening the correspondence to classical linear logic in exchange for a more expressive language, as in Priority CP (PCP)~\cite{dardhagay18}. PCP decouples CP's cut rule into two separate constructs: a construct for parallel composition, by introducing a mix rule; and a construct for restriction, by introducing a cycle rule. Differently from previous work, PCP allows cyclic structures and restores deadlock freedom by adding priorities to types following Kobayashi~\cite{kobayashi06} and Padovani~\cite{padovani14}. Priorities establish an order of actions in a $\pi$-calculus process and are used to rule out bad interleaving of channels, guaranteeing deadlock freedom.

With regards to GV, the original work on session types in GV~\cite{gayvasconcelos10,gayvasconcelos12} did not satisfy deadlock freedom. Later on, calculi in the GV family~\cite{wadler14,lindleymorris15} have achieved deadlock freedom via a syntactic restriction, \ie by combining channel creation and thread spawning into a single operation, called fork. The fork construct is the term representation of the cut construct in CP, which as mentioned above, evaluates to a channel restriction combined with a parallel composition. As with CP, this is restrictive because it limits communication structures to only trees.

>>>>

\begin{document}
\section{Introduction}
Session types have been studied for many programming paradigms, including concurrent and functional paradigms. In particular, \emph{binary} session types have been defined for the $\pi$-calculus~\cite{sangiorgiwalker01}---a process calculus for communication and concurrency---originally by Honda~\etal~\cite{honda93,takeuchihonda94,hondavasconcelos98}; and later on for Good Variation (GV)~\cite{wadler15,lindleymorris15}---a linear concurrent $\lambda$-calculus.
Binary session types guarantee communication safety and session fidelity, namely that communication proceeds without any type mismatch and it follows the predefined session protocol.
Another key property in a concurrent setting is \emph{deadlock freedom}, stating that agents communicate without getting stuck in cyclic dependencies. Deadlock freedom is a more involving property and binary session types \emph{alone} are not enough to rule out deadlocks in cyclic structures of different interleaved sessions. Several techniques have been developed to address deadlock freedom in the $\pi$-calculus and concurrent $\lambda$-calculus with session types, as detailed below.

With regards to $\pi$-calculus, a growing line of work on deadlock freedom leverages the Curry-Howard correspondences between intuitionistic or classical linear logic and the $\pi$-calculus with binary session types~\cite{cairespfenning10,wadler14}, which guarantees deadlock freedom by restricting the allowed processes to only trees. This is achieved by merging constructs for parallel composition and channel restriction under the logical cut rule, which forces processes to share \emph{only one} channel for communication, thus eliminating cyclic dependencies and deadlocks. Recent developments in the Classical Processes (CP)~\cite{wadler14} line of work have led to various approaches to decoupling these constructs, either by maintaining a strong correspondence to classical linear logic, as in Hypersequent CP~\cite{kokkemontesi19popl,kokkemontesi19tlla}, or by weakening the correspondence to classical linear logic in exchange for a more expressive language, as in Priority CP (PCP)~\cite{dardhagay18}. PCP decouples CP's cut rule into two separate constructs: a construct for parallel composition, by introducing a mix rule; and a construct for restriction, by introducing a cycle rule. Differently from previous work, PCP allows cyclic structures and restores deadlock freedom by adding priorities to types following Kobayashi~\cite{kobayashi06} and Padovani~\cite{padovani14}. Priorities establish an order of actions in a $\pi$-calculus process and are used to rule out bad interleaving of channels. Dardha and Gay prove cycle-elimination of PCP (following the standard cut-elimination proof) by leveraging priorities. Then, deadlock freedom follows as a result of the cycle-elimination proof.
%
With regards to GV, the original work on session types in GV~\cite{gayvasconcelos12} did not satisfy deadlock freedom. Later on, calculi in the GV family~\cite{wadler15,lindleymorris15} have achieved deadlock freedom via a syntactic restriction, \ie by combining channel creation and thread spawning into a single operation, called fork. The fork construct is the term representation of the cut construct in CP, which as mentioned above, evaluates to a channel restriction combined with a parallel composition. As with CP, this is restrictive because it limits process structures to only trees.
